---
title: "VTE validation"
author: "Yeji Ko"
output:
  rmdformats::robobook:
    thumbnails: no
    lightbox: yes
    gallery: no
    highlight: tango
    use_bookdown: no
    toc_depth: 2
    fig_caption: yes
    code_folding: hide
    embed_fonts: no
    keep_md: no
    number_sections: yes
    html_document:
    df_print: paged
---

```{r, warning=FALSE}
library(dplyr)
library(tidyr)
library(mice)
library(rms)
library(stringr)
library(condMVNorm)
library(knitr)

setwd("/Users/yejiko/Downloads/PhD")
source("/Users/yejiko/Downloads/PhD/Grant/prelim/function.R")

# data from clarity: used in simulation study
# clarity = readRDS("/Users/yejiko/Downloads/PhD/Grant/clarity.analysis.dataset.rds")

# updated data from clarity: used in pruned model
clarity_new = readRDS("/Users/yejiko/Downloads/clarity.data.formatted.2025.09.22.rds")
```

# Model development in the derivation cohort

We developed a pruned logistic regression model with 17 predictors for HA-VTE using the derivation cohort from clarity data. Missing values in 5 predictors were imputed with a fixed median value of each predictor. 

```{r, warning=FALSE}
clarity_new$heart.rate.trunc <- ifelse(clarity_new$HEART_RATE_FIRST_VALUE >= 144, 144, clarity_new$HEART_RATE_FIRST_VALUE)
clarity_new = clarity_new %>% mutate(admit.type = factor(str_trim(as.character(admit.type))))
data.dev <- clarity_new[clarity_new$validation.starting.cy2021 == "Build",]
data.dev = data.dev %>% dplyr::select(vte.hac, admit.type, CENTRAL_LINE_DURING_ENC, HYPOXEMIA, ECM_WGHTLOSS, ECM_COAG,
                                       CRP_FIRST_VAL, BUN_FIRST_VAL, ECM_RENLFL_MOD, ECM_RENLFL_SEV, heart.rate.trunc,
                                       ECM_PSYCHOSES, SODIUM_FIRST_VAL, CHLORIDE_FIRST_VAL, ECM_PARALYSIS, ECM_ANEMDEF,
                                       ECM_ULCER_PEPTIC, ECM_LUNG_CHRONIC)
# Median Value Imputation
# Step 1: Calculate median values from the Development cohort
median.values <- data.dev %>%
  summarise(across(where(is.numeric), median, na.rm = TRUE))

# Step 2: Apply median imputation using the Development cohort medians
data.dev <- data.dev %>%
  mutate(across(where(is.numeric), 
                ~ ifelse(is.na(.), median.values[[cur_column()]], .)))

mod.prev.vars <- lrm(vte.hac ~ CENTRAL_LINE_DURING_ENC +
                      #HEART_RATE_FIRST_VALUE +
                       admit.type +
                       HYPOXEMIA +
                       #THROMBOSIS_HX +
                       ECM_WGHTLOSS +
                       ECM_COAG +
                       CRP_FIRST_VAL +
                       BUN_FIRST_VAL +
                       ECM_RENLFL_MOD +
                       ECM_RENLFL_SEV +
                       heart.rate.trunc +
                       ECM_PSYCHOSES +
                       SODIUM_FIRST_VAL +
                       CHLORIDE_FIRST_VAL+
                       ECM_PARALYSIS +
                       ECM_ANEMDEF+
                       ECM_ULCER_PEPTIC+
                       ECM_LUNG_CHRONIC, data = data.dev)

print(mod.prev.vars)
```

# Development cohort performance

```{r, warning=FALSE}
data.dev$pred.prev.mod <- predict(mod.prev.vars, data.dev, type = "fitted")
data.dev$vte.hac.numeric <- as.numeric(as.character(data.dev$vte.hac))
val.prob(data.dev$pred.prev.mod, data.dev$vte.hac.numeric) # 0.826
```

# Validation cohort exploration

```{r, warning=FALSE}
data.val <- clarity_new[clarity_new$validation.starting.cy2021 == "Validation",]
data.val = data.val %>% dplyr::select(vte.hac, admit.type, CENTRAL_LINE_DURING_ENC, HYPOXEMIA, ECM_WGHTLOSS, ECM_COAG,
                                       CRP_FIRST_VAL, BUN_FIRST_VAL, ECM_RENLFL_MOD, ECM_RENLFL_SEV, heart.rate.trunc,
                                       ECM_PSYCHOSES, SODIUM_FIRST_VAL, CHLORIDE_FIRST_VAL, ECM_PARALYSIS, ECM_ANEMDEF,
                                       ECM_ULCER_PEPTIC, ECM_LUNG_CHRONIC)

100*colSums(is.na(data.val))/nrow(data.val)
```

We have 5 variables (all continuous) with missing values: CRP_FIRST_VAL, BUN_FIRST_VAL, heart.rate.trunc, SODIUM_FIRST_VAL, CHLORIDE_FIRST_VAL. CRP and BUN follow non-normal distribution, whereas heart rate, sodium, and chloride level are normally distributed. After log transformation, BUN follow normal distribution, confirming that its original distribution is log-normal distribution. On the contrary, CRP does not show normal distribution after log transformation.

```{r, warning=FALSE}
par(mfrow = c(1,2))
hist(data.val$CRP_FIRST_VAL, main="CRP distribution") # log normal
hist(log(data.val$CRP_FIRST_VAL), maub="log transformed CRP distribution")
par(mfrow = c(1,2))
hist(data.val$BUN_FIRST_VAL, main="BUN distribution") # log normal
hist(log(data.val$BUN_FIRST_VAL), main="log transformed BUN distribution")
par(mfrow = c(1,3))
hist(data.val$heart.rate.trunc)
hist(data.val$SODIUM_FIRST_VAL)
hist(data.val$CHLORIDE_FIRST_VAL)
```

# Validation cohort performance after imputation

## 1. Fixed value with median

```{r, warning=FALSE}
data.val <- clarity_new[clarity_new$validation.starting.cy2021 == "Validation",]
data.val = data.val %>% dplyr::select(vte.hac, admit.type, CENTRAL_LINE_DURING_ENC, HYPOXEMIA, ECM_WGHTLOSS, ECM_COAG,
                                       CRP_FIRST_VAL, BUN_FIRST_VAL, ECM_RENLFL_MOD, ECM_RENLFL_SEV, heart.rate.trunc,
                                       ECM_PSYCHOSES, SODIUM_FIRST_VAL, CHLORIDE_FIRST_VAL, ECM_PARALYSIS, ECM_ANEMDEF,
                                       ECM_ULCER_PEPTIC, ECM_LUNG_CHRONIC)


data_imp <- data.val %>%
  mutate(across(where(is.numeric), 
                ~ ifelse(is.na(.), median.values[[cur_column()]], .)))

med = data_imp$pred.prev.mod = predict(mod.prev.vars, data_imp, type = "fitted") 
data_imp$vte.hac.numeric <- as.numeric(as.character(data_imp$vte.hac))
val.prob(data_imp$pred.prev.mod, data_imp$vte.hac.numeric) # 0.827
```

## 2. Conditional modeling

```{r, warning=FALSE}
mice_obj <- mice(data = data.val[, 2:18], m = 1, method = 'pmm', print = FALSE)
da_imp = complete(mice_obj, 1) # single imputation instead of multiple imputation

hist(data_imp$CRP_FIRST_VAL, main = "Distribution of CRP after imputation")
cmi = data_imp$pred.prev.mod <- predict(mod.prev.vars, da_imp, type = "fitted")
val.prob(data_imp$pred.prev.mod, data_imp$vte.hac.numeric) # 0.824
```

## 3. Joint modeling

```{r, warning=FALSE}
convert_binary_to_factor <- function(data) {
  
  data[] <- lapply(data, function(x) {
    
    # Identify binary numeric variables (0/1 only)
    if (is.numeric(x) && all(na.omit(x) %in% c(0, 1))) {
      return(factor(x, levels = c(0, 1), labels = c("0", "1")))
    }
    
    # Identify logical variables
    if (is.logical(x)) {
      return(factor(x, levels = c(FALSE, TRUE), labels = c("0", "1")))
    }
    
    # Otherwise keep original variable
    return(x)
  })
  
  return(as.data.frame(data))
}


data.val.wide <- data.val[,2:18] %>%
    fastDummies::dummy_cols(select_columns = "admit.type",
                            remove_selected_columns = TRUE,
                            remove_first_dummy = FALSE) %>%
    dplyr::rename_with(~ gsub("admit.type_", "", .x)) %>%
    mutate(dplyr::across(where(is.factor), ~ as.numeric(as.character(.))))
  
mean <- colMeans(data.val.wide, na.rm = TRUE)
var <- cov(data.val.wide, use = "complete.obs")
var[which(var != diag(var))] <- 0
cov <- var + diag(ncol(var)) * 1e-5
  
m_times = 80
jm_imps <- joint.MI(data.val.wide, mean, cov, n.imp = m_times)
jm_list <- lapply(1:m_times, function(m){
    tmp <- jm_imps[, , m]
    cols <- c("Elective", "Emergency", "Trauma", "Urgent", "Missing")
    tmp  %<>% data.frame() %>%
                pivot_longer(cols = all_of(cols),
                names_to = "admit.type",
                values_to = "value") %>%
                filter(value == 1) %>% dplyr::select(-value) %>%
                convert_binary_to_factor()
  })

# Helper: average predictions across multiple imputations
average_preds <- function(model, imp_list){
  preds <- sapply(1:length(imp_list), function(m){
    newdata <- imp_list[[m]]
    predict(model, newdata=newdata, type="fitted")
    })
  rowMeans(preds)
}

hist(jm_list[[1]]$CRP_FIRST_VAL, main = "Distribution of CRP after imputation")
jmi = data_imp$pred.prev.mod = average_preds(mod.prev.vars, jm_list)
val.prob(data_imp$pred.prev.mod, data_imp$vte.hac.numeric) # 0.818
```

## 4. Data augmentation

```{r, warning=FALSE}
# CRP and BUN are transformed
da_imps <- replicate(
  m_times,
  data_augment_em_mult(
    data.val.wide,
    max_iter = 10,
    verbose = FALSE
  ),
  simplify = FALSE
)

# check what one run looks like
str(da_imps[[1]])

# convert each run to a numeric matrix and stack into 3D array
da_imps_arr <- simplify2array(lapply(da_imps, function(x) as.matrix(x)))

dim(da_imps_arr)
# should be: 171891 21 m_times

# example: 1st imputed dataset (as matrix)
da_imps_arr[,,1]

da_list <- lapply(1:m_times, function(m){
    tmp <- da_imps_arr[, , m]
    cols <- c("Elective", "Emergency", "Trauma", "Urgent", "Missing")
    tmp  %<>% data.frame() %>%
                pivot_longer(cols = all_of(cols),
                names_to = "admit.type",
                values_to = "value") %>%
                filter(value == 1) %>% dplyr::select(-value) %>%
                convert_binary_to_factor()
  })

hist(da_imps$CRP_FIRST_VAL, main = "Distribution of CRP after imputation")
hist(da_imps$BUN_FIRST_VAL, main = "Distribution of heart rate after imputation")

# use average_preds instead of predict()
da = data_imp$pred.prev.mod =  average_preds(mod.prev.vars, da_list)
val.prob(data_imp$pred.prev.mod, data_imp$vte.hac.numeric) # 0.823
```

## 5. MICE

```{r, warning=FALSE}
mice_obj <- mice(data = data.val[, 2:18], m = m_times, method = 'pmm', print = FALSE)
mice_list <- lapply(1:mice_obj$m, function(m){
    tmp <- complete(mice_obj, m)
    tmp
  })

hist(data_imp$CRP_FIRST_VAL, main = "Distribution of CRP after imputation")
mice = data_imp$pred.prev.mod = average_preds(mod.prev.vars, mice_list)
val.prob(data_imp$pred.prev.mod, data_imp$vte.hac.numeric) # 0.824
```

# Summary

When applying different imputation methods to the validation cohort to evaluate the pruned model, MICE yielded the best discrimination and calibration performance. Median imputation achieved comparable discrimination (C statistic), likely because it aligns with the imputation strategy used during model development, but its calibration was notably inferior to MICE. Although data augmentation produced reasonable distributions for imputed variables (see histograms above), it performed very poorly, particularly in calibration.

```{r, warning=FALSE}
par(mfrow = c(3,2)) 
v_med = val.prob(med, data_imp$vte.hac.numeric)
v_cmi = val.prob(cmi, data_imp$vte.hac.numeric)
v_jmi = val.prob(jmi, data_imp$vte.hac.numeric)
v_da = val.prob(da, data_imp$vte.hac.numeric)
v_mice = val.prob(mice, data_imp$vte.hac.numeric)

# c statistics, calibration slope and intercept
results_table <- data.frame(
  Imputation_Method = c(
    "Median Imputation",
    "Conditional Modeling Imputation",
    "Joint Modeling Imputation",
    "Data Augmentation",
    "MICE"),
  AUC = c(0.827, 0.800, 0.818, 0.822, 0.824), 
  cal_slope = c(0.995, 0.896, 1.012, 0.931, 0.953),
  cal_intercept = c(0.281, 0.396, 0.220, -0.111, -0.037))

names(results_table)[2:4] = c("C statistic", "Calibration slope", "Calibration intercept")
kable(results_table)
```



